Relative parabolic factorizations of Chevalley groups

### 2015-08-20 paper.tex
Small typography and grammar/punctuation fixes

\section{Introduction}\label{sec:intro}
Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system with a fixed basis of simple roots $\Pi$.
Denote by $G=\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over $R$ and by $\E(\Phi, R)$ its {\it elementary subgroup}, 
 i.\,e. the subgroup generated by root unipotents $t_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ denote by $\G(\Phi, R, I)$ the {\it relative Chevalley group} (also called {\it principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item {\it Gauss decomposition} i.\,e. decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item {\it Bass---Kolster} decomposition $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item {\it Dennis---Vaserstein} decomposition $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $t_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

%TODO: Illustrate each statement of the text below with a reference:
There are two main reasons to study parabolic factorizations.
Recall that a number of various factorizations of $\G$ are known over small-dimensional rings.
For example, Bruhat decomposition is the key instrument in the study of linear groups over fields, whereas Gauss decomposition and unitriangular factorizations are useful tools in the study of groups over local rings.
Moreover, for each particular analytic or arithmetic setting there is a special decomposition theorem (Iwasawa, Cartan and Iwahori decompositions, to mention a few).
Unfortunately, neither of these decompositions can be carried over to rings of higher dimension.
The main advantage of parabolic factorizations lies in the fact that they remain true for a broader class of rings described in terms of milder ring-theoretic assumptions formulated in terms of the {\it stable rank} of the ground ring.
Their another advantage is that they prove useful in the study of asymptotical properties of Chevalley groups. We are going to discuss this theme in more detail below.

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the {\it width of $G$ with respect to $X$} i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $t_\alpha(\xi)$ has been extensively studied in a number of works (see e.\,g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\,g. localizations of Dedeking rings of arithmetic type.
For example, W.~van der Kallen has shown that the width of $\SL_3(R)$ is infinite already for a euclidean ring $R=\mathbb{C}[t]$.

@@

\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $w(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk\Phi-1$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $w(\E(\Phi,R,I),\mathcal{Z})\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $w(\G(\Phi,\mathcal{O}_S,I),\mathcal{Z})$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}
%In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
%Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).
...
\end{proof}
The particular case $\Phi = \rA_l$, $R = \mathbb{Z}$ of the above theorem was considered by U.~Hadad in~\cite[Theorem~1.6]{Ha12}.

The second application concerns subsystem factorizations of Chevalley groups formulated in terms of submaximal subgroups of type $\rA_\ell$.
The case when $R=I$ is a field this question was studied by N.~Nikolov in \cite{NikProdDecomp}. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $n$ over a finite field.
Then $G$ equals the product of at most $200$ conjugates of $\SL_n$. \end{thm*}

It is also classically known that over arbitrary field $F$ the group $\SL(n+1, F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, from the classical Dennis---Vaserstein theorem (see~\cite{Va69} and \cite[Lemma~2.1]{ST76})
it follows that $\SL_{n+1}(R, I)$ is a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leq n-1$.
Using another Dennis---Vaserstein type factorization we obtain the following result.
\begin{thm} Assume that $\sr(I) = 2$ then the group $\Spin_{2\ell}(R, I)=\G_{sc}(\rD_\ell, R, I)$ is a product of at most $C$ copies of subgroups of type $\rA_{\ell-1}$. \end{thm}

>>>>

\section{Introduction}\label{sec:intro}
Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system with a fixed basis of simple roots $\Pi$.
Denote by $G=\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over $R$ and by $\E(\Phi, R)$ its \emph{elementary subgroup}, 
i.\,e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (also called \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item \emph{Bass---Kolster} decomposition $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item \emph{Dennis---Vaserstein} decomposition $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $t_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

%TODO: Illustrate each statement of the text below with a reference:
There are two main reasons to study parabolic factorizations.
Recall that a number of various factorizations of $\G$ are known over small-dimensional rings.
For example, Bruhat decomposition is the key instrument in the study of linear groups over fields, whereas Gauss decomposition and unitriangular factorizations are useful tools in the study of groups over local rings.
Moreover, for each particular analytic or arithmetic setting there is a special decomposition theorem (Iwasawa, Cartan and Iwahori decompositions, to mention a few).
Unfortunately, neither of these decompositions can be carried over to rings of higher dimension.
The main advantage of parabolic factorizations lies in the fact that they remain true for a broader class of rings described in terms of milder ring-theoretic assumptions formulated in terms of the \emph{stable rank} of the ground ring.
Their another advantage is that they prove useful in the study of asymptotical properties of Chevalley groups. We are going to discuss this theme in more detail below.

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $x_\alpha(\xi)$ has been extensively studied in a number of works (see e.\,g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\,g. localizations of Dedeking rings of arithmetic type.
For example, W.~van der Kallen has shown that the width of $\SL_3(R)$ is infinite already for a Euclidean ring $R=\mathbb{C}[t]$.

@@

\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk\Phi-1$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $W(\E(\Phi,R,I),\mathcal{Z})\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I),\mathcal{Z})$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}
%In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
%Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).
...
\end{proof}
The particular case $\Phi = \rA_\ell$, $R = \mathbb{Z}$ of the above theorem was considered by U.~Hadad in~\cite[Theorem~1.6]{Ha12}.

The second application concerns subsystem factorizations of Chevalley groups formulated in terms of submaximal subgroups of type $\rA_\ell$.
In case when $R=I$ is a field this question was studied by N.~Nikolov in \cite{NikProdDecomp}. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $n$ over a finite field.
Then $G$ equals the product of at most $200$ conjugates of $\SL_n$. \end{thm*}

It is also classically known that over an arbitrary field $F$ the group $\SL(n+1, F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, from the classical Dennis---Vaserstein theorem (see~\cite{Va69} and \cite[Lemma~2.1]{ST76})
it follows that $\SL_{n+1}(R, I)$ is a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.
Using another Dennis---Vaserstein type factorization we obtain the following result.
\begin{thm} Assume that $\sr(I) = 2$ then the group $\Spin_{2\ell}(R, I)=\G_{sc}(\rD_\ell, R, I)$ is a product of at most $C$ copies of subgroups of type $\rA_{\ell-1}$. \end{thm}
